Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-05 _tex/index.tex
Built site for gh-pages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    causal inference \sep probability \sep Thurstone \sep comparative
judgement \sep directed acyclic graph \sep structural causal
models \sep 
    statistical modeling
\end{keyword}
\end{frontmatter}

@@

{In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher trait level. After conducting multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Additionally, research on validity suggests that scores generated by CJ
can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023},
while research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aim to address all three concerns simultaneously.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The literature indicates that the scores have been used to identify
`misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power, increasing
the risk of committing a type I error \citep{McElreath_2020}, which is
when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

@@

@@

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={Probability, Directed Acyclic Graphs, Bayesian
methods, Thurstonian model, Comparative judgement, Structural Causal
Models, Statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Probability \sep Directed Acyclic Graphs \sep Bayesian
methods \sep Thurstonian model \sep Comparative
judgement \sep Structural Causal Models \sep 
    Statistical modeling
\end{keyword}
\end{frontmatter}

@@

In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher trait level. After conducting multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts, where quality describes the underlying trait of interest
and the texts serve as the stimuli
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Additionally, research on validity suggests that scores generated by CJ
can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of CJ studies, the literature's
unsystematic and fragmented research approaches have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aims to address all three concerns simultaneously.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The scores have been used to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power, increasing
the risk of committing a type I error \citep{McElreath_2020}. A type I
error results when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

@@

From early on in the literature, it has been clear that the BTL model
represents a statistical articulation of Thurstone's Case 5
\citeyearpar{Thurstone_1927}. Talk about \citet{Pollitt_et_al_2003} and
\citet{Bramley_2008}.

What case 5 implies, Assumptions. Not a normal distribution but a
logistic distribution

Table with assumptions

@@

\subsection{Appendix A: Ignoring uncertainty}\label{sec-appendix-A}

\subsection{Appendix B: The five cases of
Thurstone}\label{sec-appendix-B}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}
